Ownership types in theory and practice (in Rust)

### 2023-01-22 paper/paper.tex
Added rust and cpp syntax highlighting to paper and slides

\usepackage{booktabs} % For formal tables
\usepackage{subcaption}
\usepackage{tikz}
\usetikzlibrary{fit}
\usepackage{pgfplots}
\usepackage{pgfplotstable}
\usepackage{hyphenat}
\usepackage{todonotes}
\usepackage[babel]{csquotes}
\usepackage{listings}
\lstset{frame=tb,
  % language=Rust,
  aboveskip=3mm,
  belowskip=3mm,
  % showstringspaces=false,
  columns=flexible,
  basicstyle={\small\ttfamily},
  numbers=left,
  numbersep=5pt,
  xleftmargin=\parindent,
  numberstyle=\color{gray},
  % keywordstyle=\color{blue},
  % commentstyle=\color{dkgreen},
  % stringstyle=\color{mauve},
  breaklines=true,
  % breakatwhitespace=true,
  tabsize=2
}
\pagenumbering{gobble}
\definecolor{cgreen}{RGB}{60,201,65}

@@

\begin{lstlisting}
std::vector<int> v { 10, 11 };
int *vptr = &v[1]; // points into v
v.push_back(12);
std::cout << *vptr; // bug (use-after-free)
\end{lstlisting}

@@

\begin{lstlisting}
let mut v = vec![10, 11];
let vptr = &mut v[1]; // points *into* v
v.push(12);
println!("{}", *vptr); // compiler error
\end{lstlisting}

@@

\begin{lstlisting}
let a = HeapBasedType::default();
let b = a;
\end{lstlisting}
In Rust terminology, \emph{a} is \emph{moved} into \emph{b}.
After the move, \emph{a} is no longer valid, and \emph{b} is the new owner of the value.

Heap-based values will always be generated by a function at runtime.
In the above example, the return value of the function \verb|HeapBasedType::default()| is moved into \emph{a}.
Furthermore, one can generalize the above scenario to ownership being transferred from the variable \emph{a} to the variable \emph{b} when \emph{b} is assigned to the return value \emph{a} of \emph{foo}.
\begin{lstlisting}
fn main() {
  let b = foo();
}
fn foo() {
  let a = HeapBasedType::default();
  a
}
\end{lstlisting}

Lastly, ownership is transferred from the variable \emph{a} to the variable \emph{b} when \emph{a} is passed to function \emph{foo}, which takes a parameter \emph{b}.
\begin{lstlisting}
fn main() {
  let a = HeapBasedType::default();
  foo(a);
}
fn foo(b: HeapBasedType) {
  // do something with b
}
\end{lstlisting}
In \emph{foo}, \emph{b} is the owner of the value until it is \emph{dropped} at the end of the function.
Accessing the previous owner of a value after the ownership of that value has been transferred elsewhere will lead to a compile-time error.
In this case, trying to access \emph{a} after line 3 would not compile, as \emph{a} is no longer the owner of the value.

The heap-allocated \emph{String} type in Rust provides a good example demonstrating ownership transfer of heap-based values.
\begin{lstlisting}
let x: String = String::from("hello world");
let y: String = x;
println!("x: {}, y: {}", x, y);
\end{lstlisting}
After \emph{x} is moved into \emph{y} in line 2, \emph{x} is no longer valid.
Therefore, Rust prevents you from accessing the invalid \emph{x} in the third line, and the program will not compile.

@@

The following piece of code \cite{rust-book}, which is similar to the code from above but replaces the \emph{String} type with string literals, demonstrates how stack-based string literals are \emph{copied} instead of \emph{moved} into other variables in Rust.
\begin{lstlisting}
let x: &str = "hello world";
let y: &str = x;
println!("x: {}, y: {}", x, y);
\end{lstlisting}
The value of \emph{x} is copied into \emph{y}, after which both variables are still valid.
Therefore, the program compiles and "x: hello world, y: hello world" is printed to stdout.

@@

\begin{lstlisting}
struct Car {
  engine: Engine,
  wheels: Wheels,
}
fn main() {
  let car = Car::default();
  let moved_engine = car.engine; // valid
  car.drive() // error: use of partially moved value: 'car'
}
\end{lstlisting}

@@

The following adjustment to the code snippet from above creates a deep copy of \emph{x} and compiles successfully.
\begin{lstlisting}
let x: String = String::from("hello world");
let y: String = x.clone();
println!("x: {}, y: {}", x, y);
\end{lstlisting}

@@

For example, a string literal can be appended to a mutable String.
\begin{lstlisting}
let mut s = String::from("hello");
s.push_str("hello");
\end{lstlisting}

@@

\begin{lstlisting}
fn main() {
  let s = String::from("hello");
  let r1 = &s;
  let r2 = &s;
  print_str(r1);
  print_str(r2);
}
fn print_str(s: &String) {
  println!("printed from a reference: {}", s);
}
\end{lstlisting}

@@

\begin{lstlisting}
fn main() {
  let mut s = String::from("hello");
  modify_str(&mut s);
}
fn modify_str(s: &mut String) {
  s.push_str(" world");
}
\end{lstlisting}

@@

\begin{lstlisting}
let mut s = String::from("hello");

@@

% TODO: format so spaces align properly
\begin{lstlisting}
let mut v = vec![10, 11];
let vptr = &mut v[1];  // lifetime 'a starts
Vec::push(&mut v, 12); // lifetime 'b
println!("{}", *vptr); // lifetime 'a ends
\end{lstlisting}

@@

By reordering lines 2 and 3, the lifetimes of the two mutable references will no longer overlap, allowing for successful compilation of the program.
\begin{lstlisting}
let mut v = vec![10, 11];
Vec::push(&mut v, 12); // lifetime 'b
let vptr = &mut v[1];  // lifetime 'a starts
println!("{}", *vptr); // lifetime 'a ends
\end{lstlisting}

@@

\begin{lstlisting}
S s{};
S s_copied{s}; // copy constructor
S s_moved{std::move(s)}; // move constructor
// problem: 's' can still be used
\end{lstlisting}

@@

\begin{lstlisting}
let s = S::default();
let s_copied = s.clone();
let s_moved = s; // 's' is moved
// compiler prevents use of 's'
\end{lstlisting}

>>>>

\usepackage{booktabs} % For formal tables
\usepackage{subcaption}
\usepackage{tikz}
\usetikzlibrary{fit}
\usepackage{pgfplots}
\usepackage{pgfplotstable}
\usepackage{hyphenat}
\usepackage{todonotes}
\usepackage[babel]{csquotes}
\usepackage{listings}
\usepackage{listings-rust}

\definecolor{cgreen}{RGB}{60,201,65}
\definecolor{dkgreen}{rgb}{0,0.6,0}
\definecolor{mauve}{rgb}{0.58,0,0.82}

\lstset{frame=tb,
  aboveskip=3mm,
  belowskip=3mm,
  showstringspaces=false,
  columns=flexible,
  basicstyle={\small\ttfamily},
  numbers=left,
  numbersep=5pt,
  xleftmargin=\parindent,
  numberstyle=\color{gray},
  keywordstyle=\bfseries\color{blue},
  commentstyle=\color{dkgreen},
  stringstyle=\color{mauve},
  breaklines=true,
  % breakatwhitespace=true,
  tabsize=2
}
\pagenumbering{gobble}

@@

\begin{lstlisting}[language=C++]
std::vector<int> v { 10, 11 };
int *vptr = &v[1]; // points into v
v.push_back(12);
std::cout << *vptr; // bug (use-after-free)
\end{lstlisting}

@@

\begin{lstlisting}[language=Rust]
let mut v = vec![10, 11];
let vptr = &mut v[1]; // points *into* v
v.push(12);
println!("{}", *vptr); // compiler error
\end{lstlisting}

@@

\begin{lstlisting}[language=Rust]
let a = HeapBasedType::default();
let b = a;
\end{lstlisting}
In Rust terminology, \emph{a} is \emph{moved} into \emph{b}.
After the move, \emph{a} is no longer valid, and \emph{b} is the new owner of the value.

Heap-based values will always be generated by a function at runtime.
In the above example, the return value of the function \verb|HeapBasedType::default()| is moved into \emph{a}.
Furthermore, one can generalize the above scenario to ownership being transferred from the variable \emph{a} to the variable \emph{b} when \emph{b} is assigned to the return value \emph{a} of \emph{foo}.
\begin{lstlisting}[language=Rust]
fn main() {
  let b = foo();
}
fn foo() {
  let a = HeapBasedType::default();
  a
}
\end{lstlisting}

Lastly, ownership is transferred from the variable \emph{a} to the variable \emph{b} when \emph{a} is passed to function \emph{foo}, which takes a parameter \emph{b}.
\begin{lstlisting}[language=Rust]
fn main() {
  let a = HeapBasedType::default();
  foo(a);
}
fn foo(b: HeapBasedType) {
  // do something with b
}
\end{lstlisting}
In \emph{foo}, \emph{b} is the owner of the value until it is \emph{dropped} at the end of the function.
Accessing the previous owner of a value after the ownership of that value has been transferred elsewhere will lead to a compile-time error.
In this case, trying to access \emph{a} after line 3 would not compile, as \emph{a} is no longer the owner of the value.

The heap-allocated \emph{String} type in Rust provides a good example demonstrating ownership transfer of heap-based values.
\begin{lstlisting}[language=Rust]
let x: String = String::from("hello world");
let y: String = x;
println!("x: {}, y: {}", x, y);
\end{lstlisting}
After \emph{x} is moved into \emph{y} in line 2, \emph{x} is no longer valid.
Therefore, Rust prevents you from accessing the invalid \emph{x} in the third line, and the program will not compile.

@@

The following piece of code \cite{rust-book}, which is similar to the code from above but replaces the \emph{String} type with string literals, demonstrates how stack-based string literals are \emph{copied} instead of \emph{moved} into other variables in Rust.
\begin{lstlisting}[language=Rust]
let x: &str = "hello world";
let y: &str = x;
println!("x: {}, y: {}", x, y);
\end{lstlisting}
The value of \emph{x} is copied into \emph{y}, after which both variables are still valid.
Therefore, the program compiles and "x: hello world, y: hello world" is printed to stdout.

@@

\begin{lstlisting}[language=Rust]
struct Car {
  engine: Engine,
  wheels: Wheels,
}
fn main() {
  let car = Car::default();
  let moved_engine = car.engine; // valid
  car.drive() // error: use of partially moved value: 'car'
}
\end{lstlisting}

@@

The following adjustment to the code snippet from above creates a deep copy of \emph{x} and compiles successfully.
\begin{lstlisting}[language=Rust]
let x: String = String::from("hello world");
let y: String = x.clone();
println!("x: {}, y: {}", x, y);
\end{lstlisting}

@@

For example, a string literal can be appended to a mutable String.
\begin{lstlisting}[language=Rust]
let mut s = String::from("hello");
s.push_str("hello");
\end{lstlisting}

@@

\begin{lstlisting}[language=Rust]
fn main() {
  let s = String::from("hello");
  let r1 = &s;
  let r2 = &s;
  print_str(r1);
  print_str(r2);
}
fn print_str(s: &String) {
  println!("printed from a reference: {}", s);
}
\end{lstlisting}

@@

\begin{lstlisting}[language=Rust]
fn main() {
  let mut s = String::from("hello");
  modify_str(&mut s);
}
fn modify_str(s: &mut String) {
  s.push_str(" world");
}
\end{lstlisting}

@@

\begin{lstlisting}[language=Rust]
let mut s = String::from("hello");

@@

% TODO: format so spaces align properly
\begin{lstlisting}[language=Rust]
let mut v = vec![10, 11];
let vptr = &mut v[1];  // lifetime 'a starts
Vec::push(&mut v, 12); // lifetime 'b
println!("{}", *vptr); // lifetime 'a ends
\end{lstlisting}

@@

By reordering lines 2 and 3, the lifetimes of the two mutable references will no longer overlap, allowing for successful compilation of the program.
\begin{lstlisting}[language=Rust]
let mut v = vec![10, 11];
Vec::push(&mut v, 12); // lifetime 'b
let vptr = &mut v[1];  // lifetime 'a starts
println!("{}", *vptr); // lifetime 'a ends
\end{lstlisting}

@@

\begin{lstlisting}[language=C++]
S s{};
S s_copied{s}; // copy constructor
S s_moved{std::move(s)}; // move constructor
// problem: 's' can still be used
\end{lstlisting}

@@

\begin{lstlisting}[language=Rust]
let s = S::default();
let s_copied = s.clone();
let s_moved = s; // 's' is moved
// compiler prevents use of 's'
\end{lstlisting}

### 2023-01-22 slides/slides.tex
Added rust and cpp syntax highlighting to paper and slides

\usepackage[utf8]{inputenc}
\usepackage{listings}
\usepackage{tikz}
\usetikzlibrary{fit,positioning}
% \usepackage[style=alphabetic,giveninits=true,terseinits=true]{biblatex}

\definecolor{tumblue}{RGB}{0,101,189}
\definecolor{dkgreen}{rgb}{0,0.6,0}
\definecolor{mauve}{rgb}{0.58,0,0.82}
\lstset{
  basicstyle=\footnotesize\ttfamily,
  columns=fullflexible,
  keywordstyle=\bfseries\color{blue},
  commentstyle=\color{dkgreen},
  stringstyle=\color{mauve},
}

\lstset{frame=tb,
  % language=Rust,
  aboveskip=3mm,
  belowskip=3mm,
  % showstringspaces=false,
  columns=flexible,
  basicstyle={\small\ttfamily},
  numbers=left,
  numbersep=5pt,
  xleftmargin=\parindent,
  numberstyle=\color{gray},
  % keywordstyle=\color{blue},
  % commentstyle=\color{dkgreen},
  % stringstyle=\color{mauve},
  breaklines=true,
  % breakatwhitespace=true,
  tabsize=2
}
\definecolor{cgreen}{RGB}{60,201,65}

@@

C++
\begin{lstlisting}
std::vector<int> v { 10, 11 };
int *vptr = &v[1]; // points into v
v.push_back(12);
std::cout << *vptr; // bug (use-after-free)
\end{lstlisting}

Rust
\begin{lstlisting}
let mut v = vec![10, 11];
let vptr = &mut v[1]; // points *into* v
v.push(12);
println!("{}", *vptr); // compiler error
\end{lstlisting}

>>>>

\usepackage[utf8]{inputenc}
\usepackage{listings}
\usepackage{listings-rust}
\usepackage{tikz}
\usetikzlibrary{fit,positioning}
% \usepackage[style=alphabetic,giveninits=true,terseinits=true]{biblatex}

\definecolor{tumblue}{RGB}{0,101,189}
\definecolor{dkgreen}{rgb}{0,0.6,0}
\definecolor{mauve}{rgb}{0.58,0,0.82}
% \lstset{
%   frame=tb,
%   basicstyle=\footnotesize\ttfamily,
%   columns=fullflexible,
%   numbers=left,
%   numbersep=5pt,
%   xleftmargin=\parindent,
%   keywordstyle=\bfseries\color{blue},
%   commentstyle=\color{dkgreen},
%   stringstyle=\color{mauve},
% }

\lstset{
  frame=tb,
  aboveskip=3mm,
  belowskip=3mm,
  showstringspaces=false,
  columns=flexible,
  basicstyle=\footnotesize\ttfamily,
  % basicstyle={\small\ttfamily},
  numbers=left,
  numbersep=5pt,
  % xleftmargin=\parindent,
  numberstyle=\color{gray},
  keywordstyle=\bfseries\color{blue},
  commentstyle=\color{dkgreen},
  stringstyle=\color{mauve},
  breaklines=true,
  breakatwhitespace=false,
  tabsize=2
}
\definecolor{cgreen}{RGB}{60,201,65}

@@

C++
\begin{lstlisting}[language=C++]
std::vector<int> v { 10, 11 };
int *vptr = &v[1];  // points into v
v.push_back(12);
std::cout << *vptr; // bug (use-after-free)
\end{lstlisting}

Rust
\begin{lstlisting}[language=Rust]
let mut v = vec![10, 11];
let vptr = &mut v[1];  // points *into* v
v.push(12);
println!("{}", *vptr); // compiler error
\end{lstlisting}
